Hill-Robertson Interference Causes Reduced Genetic Diversity on a Young Plant Y-chromosome

### 2016-12-20 manuscript/manuscript_Dec19.tex
fixed figure 1, new edits up to results

\section*{Materials and Methods}
\subsection*{Population Samples and Sex-Linked Genes}
We analyzed sex-linked and autosomal genes identified from Illumina RNA sequence data from 12 population samples (1 male and 1 female from each of 6 populations). Samples were collected in 2010 from throughout the native range of \textit{R. hastatulus} (locations in Table S1), and plants were grown in the glasshouse at the University of Toronto from seeds collected from open-pollinated females. We extracted RNA from leaf tissue using Spectrum Plant Total RNA kits (Sigma-Aldrich). Isolation of mRNA and cDNA synthesis was conducted according to standard Illumina RNAseq procedures, and sequencing was conducted on two Illumina HiSeq lanes with 150-bp end reads at the Genome Quebec Innovation Center. Reads from these samples were mapped to the \textit{R. hastatulus} reference transcriptome \citep{hough2014} using the Burrows–Wheeler Aligner \citep{li2010fast}, followed by Stampy \citep{lunter2011stampy}. We used Picard tools (http://picard.sourceforge.net) to process mapping alignments for the Genome Analysis Toolkit \citep{mckenna2010genome} variant calling software, and subsequently removed genes with low coverage (<10x) and low Phred Quality Scores <20. The RNAseq data from these population samples were previously reported in Hough et al. (2014), where they were used to validate the ascertainment of sex-linked genes identified through segregation analysis, and raw sequences are available from the GenBank Short Read Archive under accession no. SRP041588. Here, to focus on sex-linked genes likely to be relatively older and closer to equilibrium, we focused on the previously described set of 460 X/Y genes for which a Y homolog was found in both races (i.e., X/Y genes where the Y copy was inferred to be on the $\textnormal{Y}_\textnormal{1}$ chromosome).

@@

\subsection*{Phasing X and Y alleles}
To estimate polymorphism for X and Y sequences separately in males, it is necessary to infer the phase of SNPs in sex-linked transcripts in males. In previous work, phasing alleles on \textit{R. hastatulus} sex chromosomes was achieved using segregation analysis from a genetic cross. Here, to phase SNPs from population samples where such segregation data was unavailable, we used HAPCUT \citep{bansal2008hapcut}, a maximum-cut based algorithm that reconstructs haplotypes using sequenced fragments (Illumina read data) from the two homologous chromosomes to output a list of phased haplotype blocks containing the SNP variants on each chromosome. Because the resulting haplotype blocks produced by HAPCUT contained SNPs that were phased relative to each other, but not designated to either the X or Y chromosome, we assigned individual variants to X or Y by independently identifying fixed X-Y differences within each haplotype block (i.e., sites where all females were homozygous, and all males were heterozygous). Identifying such fixed differences within phased haplotype blocks enabled us to then infer the correct phase (X or Y) of the polymorphisms from HAPCUT’s output. In particular, this was done by matching the phase of fixed X-Y differences with their neighboring polymorphic sites: when a fixed X-Y difference occurred in the same phased haplotype block as a polymorphic site, then the polymorphic variants in that block were assigned to either X or Y based on the known phase of the fixed difference with which they were matched. SNPs that were identified outside of phased blocks, or in blocks without fixed X-Y differences, were recorded as missing data. Finally, we filtered out SNPs with quality scores < 60, and those within a distance of 10bp or less from indels. This filtering procedure resulted in alignments of X and Y sequences for 372 sex-linked genes.

We further validated the results of HAPCUT’s allele phasing by comparing the accuracy of this method with the phasing-by-segregation method that was conducted in Hough et al. (2014). To do this, we first phased the sequence data from parents and their progeny using HAPCUT’s algorithm (using the same parameters as for the population data), and then identified cases where SNPs were inferred on the Y chromosome by HAPCUT, but where the true level of polymorphism, obtained from the genetic cross, was zero. We identified 7 \% of sex-linked genes that either had phasing errors or genotyping errors. This corresponds to a SNP error rate estimate of 1.7 x 10-4. Note that this rate is very low relative to population-based estimates of polymorphism on the X and autosomes (Table 1), and therefore should have minimal effects on our estimation of the X/A ratio. However, because this rate is high relative to the expected level of polymorphism on the Y chromosome, we conducted a further filtering step for Y-linked SNPs and identified false positive SNP calls arising from: (i) phasing errors caused by gene duplicates (more than two haplotypes), (ii) polymorphisms around indels, and (iii) genotyping errors caused by low Y-expression. This final filtering was conducted by manually checking each individual putative polymorphism on the Y chromosome using IGV \citep{robinson2011integrative}.

\subsection*{Estimating nucleotide diversity on sex chromosomes and autosomes}
For each locus in our analysis, we calculated Watterson’s (1975) estimator of the population parameter $\theta=4N_{e}\mu$, where $N_{e}$ is the effective population size, and $\mu$ is the mutation rate \citep{watterson1975}, using a modified version of the Perl program Polymorphurama \citep{bachtrog2006}. To compare sex-linked and autosomal loci, we calculated the average value of $\theta$, weighted by the number of synonymous sites in each gene (\X Figure 2). We obtained 95\% confidence intervals for X/A and Y/A ratios by bootstrapping per gene using the BCa method \citep{efron1994} implemented in the Boot package in R \citep{canty2012boot}, and calculating X/A and Y/A on each iteration for 20000 replicates each. Bootstrapping was conducted on the final filtered set of 173 sex-linked, and 12355 autosomal genes.

Note that the lack of recombination on the Y chromosome implies that statistical assumptions about independence across loci are violated, suggesting that the true uncertainty in the Y/A estimate may be wider than implied by bootstrapping. To address this, we also used a maximum likelihood approach, implemented in a modified version of the MLHKA software \citep{wright2004hka}, to independently estimate a credibility interval for the Y/A ratio (Figure S1). Because of the thousands of genes involved, a likelihood method incorporating divergence to control for heterogeneity in mutation rate was not feasible, as this would require maximizing the likelihood estimate of the mutation rate for each locus independently. However, previous analysis of divergence data does not suggest important chromosomal differences in synonymous divergence contributing to diversity heterogeneity (see Results). Therefore, we assumed no heterogeneity in mutation rate, no recombination between Y-linked genes, and free recombination between autosomal loci. Our model thus had two parameters: $\theta_{autosomal}$ and \textit{f}, the ratio of effective population size of the Y chromosome to autosomes. We varied both parameters and evaluated the likelihood for \textit{f} from 0.001 to 1, and $\theta_{autosomal}$ per base pair from 0.001 to 0.01. In particular, we evaluated the likelihood of $\theta_{autosomal}$ and \textit{f}, given the observed number of synonymous segregating sites for each \textit{i} of \textit{n} autosomal loci ($S_{i}$), the number of segregating sites on the Y chromosome ($S_{y}$), and the total number of sites per locus ($L_{i}$ and $L_{y}$). The likelihood of ($\theta_{autosomal}$, \textit{f}) is therefore given by:

@@

\subsection*{Neutral predictions and the effect of sex ratio bias on diversity}
To test whether our estimated levels of diversity on X, Y and autosomal chromosomes could be explained by neutral processes, including female-biased population sex ratios or high variance in male reproductive success, we compared our normalized X/A and Y/A diversity estimates to neutral predictions for a range of values for the ratio of effective population sizes of females and males. Because we were primarily interested in determining the parameter space within which a reduced male effective population size could result in Y/A and X/A diversity ratios consistent with our data, we tested the fit of our data to predictions across the full range of $N_{e}_{f}/(N_{e}_{f} + N_{e}_{m})$ (0.1 to 0.9; we excluded all-female or all-male cases). For autosomal and sex-linked effective population sizes given by:

@@

\citep{wright1931evolution}, the corresponding expected X/A and Y/A ratios an be expressed as:

@@

respectively, where $r = N_{e}_{f}/(N_{e}_{f} + N_{e}_{m})$.  Note that when $N_{e}_{f} = N_{e}_{m}$ ($r=0.5$) then, $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$ as in the standard neutral model \citep{wright1931evolution}. With a female-biased ratio, however, the expected X/A ratio can become greater than 1, approaching 1.125 in the limit as $r\rightarrow 1$ \citep{caballero1995}.

\begin{figure}[htbp]
\centering
\noindent
\includegraphics[width=\linewidth]{figure1.jpg}
\caption{\textbf{A}. Equilibrium effective population sizes, relative to $N$, for genes on autosomes (A), X chromosomes (X), and Y chromosomes (Y) as a function of the sex ratio, expressed as $N_{e}_{f}/(N_{e}_{f} + N_{e}_{m})$. \textbf{B}. The corresponding X/A and Y/A ratios of nucleotide diversity predicted at equilibrium, assuming equal neutral mutation rates for sex-linked and autosomal genes, where $\theta=4N_{e}\mu$).
}
\end{figure}

\subsection*{Simulations of purifying selection}
To study the effects of purifying selection on expected levels of Y-chromosome diversity, we conducted forward-time simulations of haploid Y chromosomes using the software SFSCODE \citep{hernandez2008flexible}. We first estimated the distribution of fitness effects of deleterious amino acid mutations from our polymorphism data for X-linked genes using the method of \citet*{keightley2007joint}, which fits a gamma distribution of negative selection coefficients to the observed frequency distribution of nonsynonymous and synonymous polymorphisms. We then used this estimated gamma distribution to parameterize the simulations, initializing them with our estimated $\theta$ from autosomal genes, adjusted to reflect the expected neutral reduction in $N_{e}_{Y}$ for a sex ratio of $N_{e}_{f}/(N_{e}_{f} + N_{e}_{m}) = 0.6$. To match our sample size and the number of synonymous sites sampled from our data (see Supporting Information), the simulations sampled 6 haploid chromosomes, and the genome sequence contained 45,331bp of linked neutral sequence from which we calculated silent site diversity. To examine the expected reduction in diversity as a function of the number of selected sites ($L_{selected}$), we ran simulations over a range of values of $L_{selected}$, up to a maximum of 5x$10^{6}$ (Figure 3A). To estimate $L_{selected}$, we calculated the approximate likelihood of our observed data based on the proportion of simulations in which synonymous diversity, $\pi_{simulated}$, was less than 0.0001 from our empirical estimate, $\pi_{observed}$ (Figure 3B).

>>>>

\section*{Materials and Methods}
\subsection*{Population Samples and Sex-Linked Genes}
We analyzed sex-linked and autosomal genes identified from Illumina RNA sequence data from 12 population samples (1 male and 1 female from each of 6 populations). Samples were collected in 2010 from throughout the native range of \textit{R. hastatulus} (locations in Table S1), and plants were grown in the glasshouse at the University of Toronto from seeds collected from open-pollinated females. We extracted RNA from leaf tissue using Spectrum Plant Total RNA kits (Sigma-Aldrich). Isolation of mRNA and cDNA synthesis was conducted according to standard Illumina RNAseq procedures, and sequencing was conducted on two Illumina HiSeq lanes with 150-bp end reads at the Genome Quebec Innovation Center. Reads from these samples were mapped to the \textit{R. hastatulus} reference transcriptome \citep{hough2014} using the Burrows–Wheeler Aligner \citep{li2010fast}, followed by Stampy \citep{lunter2011stampy}. We used Picard tools (http://picard.sourceforge.net) to process mapping alignments for the Genome Analysis Toolkit variant calling software \citep{mckenna2010genome}, and subsequently removed genes with low coverage (<10x) and low Phred Quality Scores <20. The RNAseq data from these population samples were previously reported in \citet{hough2014}, where they were used to validate the ascertainment of sex-linked genes identified through segregation analysis, and raw sequences are available from the GenBank Short Read Archive under accession no. SRP041588. Here, to focus on sex-linked genes likely to be relatively older and closer to equilibrium, we focused on the previously described set of 460 X/Y genes for which a Y homolog was found in both races (i.e., X/Y genes where the Y copy was inferred to be on the $\textnormal{Y}_\textnormal{1}$ chromosome).

@@

\subsection*{Phasing X and Y alleles}
To estimate polymorphism for X and Y sequences separately in males, it is necessary to infer the phase of SNPs in sex-linked transcripts. In previous work, phasing alleles on \textit{R. hastatulus} sex chromosomes was achieved using segregation analysis from a genetic cross. Here, to phase SNPs from population samples where such segregation data was unavailable, we used HAPCUT \citep{bansal2008hapcut}, a maximum-cut based algorithm that reconstructs haplotypes using sequenced fragments (Illumina read data) from the two homologous chromosomes to output a list of phased haplotype blocks containing the polymorphic variants on each chromosome. Because the resulting haplotype blocks produced by HAPCUT contained SNPs that were phased relative to each other, but not designated to either the X or Y chromosome, we assigned individual variants to X or Y by independently identifying fixed X-Y differences within each haplotype block (i.e., sites where all females were homozygous, and all males were heterozygous). Identifying such fixed differences within phased haplotype blocks enabled us to then infer the correct phase (X or Y) of the polymorphisms from HAPCUT’s output. In particular, this was done by matching the phase of fixed X-Y differences with neighboring polymorphic sites: when a fixed X-Y difference occurred in the same phased haplotype block as a polymorphic site, then the variants in that block were assigned to either X or Y based on the known phase of the fixed difference with which they were matched. SNPs that were identified outside of phased blocks, or in blocks without fixed X-Y differences, were recorded as missing data. Finally, we filtered out SNPs with quality scores < 60, and those within a distance of 10bp or less from indels. This filtering procedure resulted in alignments of X and Y sequences for 372 sex-linked genes.

We further validated the results of HAPCUT’s allele phasing by comparing the accuracy of this method with the phasing-by-segregation method that was conducted in \citet{hough2014}. To do this, we first phased the sequence data from parents and their progeny using HAPCUT’s algorithm (using the same parameters as for the population data), and then identified cases where SNPs were inferred on the Y chromosome by HAPCUT, but where the true level of polymorphism, indicated by the genetic cross, was zero. We identified 7\% of sex-linked genes that either had phasing errors or genotyping errors. This corresponds to a SNP error rate estimate of 1.7 x $10^{-4}$. Note that this rate is very low relative to population-based estimates of polymorphism on the X and autosomes (Table 1), and therefore should have minimal effects on our estimation of the X/A ratio. However, because this rate is high relative to the expected level of polymorphism on the Y chromosome, we conducted a further filtering step for Y-linked SNPs and identified false positive SNP calls arising from: (i) phasing errors caused by gene duplicates (more than two haplotypes), (ii) polymorphisms around indels, and (iii) genotyping errors caused by low Y-expression. This final filtering was conducted by manually checking each individual putative polymorphism on the Y chromosome using IGV \citep{robinson2011integrative}.

\subsection*{Estimating nucleotide diversity on sex chromosomes and autosomes}
For each locus in our analysis, we calculated Watterson’s (1975) estimator of the population parameter $\theta=4N_{e}\mu$, where $N_{e}$ is the diploid effective population size, and $\mu$ is the per-base mutation rate per-generation, using a modified version of the Perl program Polymorphurama \citep{bachtrog2006}. To compare sex-linked and autosomal loci, we calculated the average value of $\theta$ for each chromosome type, weighted by the number of synonymous sites in each gene (\X Figure 2). We obtained 95\% confidence intervals for X/A and Y/A ratios by bootstrapping per gene using the BCa method \citep{efron1994} implemented in the Boot package in R \citep{canty2012boot}, and calculating X/A and Y/A on each iteration for 20000 replicates each. Bootstrapping was conducted on the final filtered set of 173 sex-linked, and 12355 autosomal genes.

Note that the lack of recombination on the Y chromosome implies that statistical assumptions about independence across loci are violated, suggesting that the true uncertainty in the Y/A estimate may be wider than implied by bootstrapping. To address this, we also used a maximum likelihood approach to independently estimate a credibility interval for the Y/A ratio (Figure S1). Because of the thousands of genes involved, a likelihood method incorporating divergence to control for heterogeneity in mutation rate was not feasible, as this would require maximizing the likelihood estimate of the mutation rate for each locus independently. However, previous analysis of divergence data does not suggest important chromosomal differences in synonymous divergence contributing to diversity heterogeneity (see Results). Therefore, we assumed no heterogeneity in mutation rate, no recombination between Y-linked genes, and free recombination between autosomal loci. Our model thus had two parameters: $\theta_{a}$, our estimate of Watterson's theta for autosomal genes, and \textit{f}, the ratio of effective population size of the Y chromosome to autosomes. We varied both parameters and evaluated the likelihood of \textit{f} from 0.001 to 1, and $\theta_{a}$ per base pair from 0.001 to 0.01. In particular, we evaluated the likelihood of $\theta_{a}$ and \textit{f}, given (1) the observed number of synonymous segregating sites for each \textit{i} of \textit{n} autosomal loci ($S_{i}$), (2) the number of segregating sites on the Y chromosome ($S_{y}$), and (3) the total number of sites per locus ($L_{i}$ and $L_{y}$). The likelihood of ($\theta_{a}$, \textit{f}) is given by:

@@

\subsection*{Neutral predictions and the effect of sex ratio bias on diversity}
To test whether our estimated levels of diversity on X, Y and autosomal chromosomes could be explained by neutral processes, including female-biased population sex ratios or high variance in male reproductive success, we compared our normalized X/A and Y/A diversity estimates to the corresponding neutral predictions for a given effective population size "sex ratio", $r=N_{e}_{f}/(N_{e}_{f} + N_{e}_{m})$. Because we were primarily interested in determining the parameter space within which a reduced male $N_{e}$ could result in Y/A and X/A diversity ratios consistent with our data, we tested the fit of our estimates to predictions across the full range of $r$ (0.1 to 0.9; we excluded all-female or all-male cases). For autosomal and sex-linked effective population sizes given by:

@@

\citep{wright1931evolution}, the corresponding expected X/A and Y/A ratios can be expressed as:

@@

respectively.  Note that when $N_{e}_{f} = N_{e}_{m}$ ($r=0.5$) then, $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$ as in the standard neutral model \citep{wright1931evolution}. With a female-biased ratio, however, the expected X/A ratio can become greater than 1, approaching 1.125 in the limit as $r\rightarrow 1$ \citep{caballero1995}.

\begin{figure}[htbp]
\centering
\noindent
\includegraphics[width=\linewidth]{figure1.jpg}
\caption{\textbf{A}. Equilibrium effective population sizes, relative to $N$, for genes on autosomes, X chromosomes, and Y chromosomes as a function of the sex ratio. \textbf{B}. The corresponding X/A and Y/A ratios of nucleotide diversity predicted at equilibrium, assuming equal neutral mutation rates for sex-linked and autosomal genes.
}
\end{figure}

\subsection*{Simulations of purifying selection}
To study the effects of purifying selection on expected levels of Y-chromosome diversity, we conducted forward-time simulations of haploid Y chromosomes using the software SFSCODE \citep{hernandez2008flexible}. We first estimated the distribution of fitness effects of deleterious amino acid mutations from our polymorphism data for X-linked genes using the method of \citet*{keightley2007joint}, which fits a gamma distribution of negative selection coefficients to the observed frequency distribution of nonsynonymous and synonymous polymorphisms. We then used this estimated gamma distribution to parameterize the simulations, initializing them with our estimated $\theta_{a}$, adjusted to reflect the neutrally expected $N_{e_{Y}}$ for a sex ratio of $r=0.6$. To match our sample size and the number of synonymous sites sampled from our data (see Supporting Information), the simulations sampled 6 haploid chromosomes, and the genome sequence contained 45,331 bp of linked neutral sequence from which we calculated silent site diversity, $\pi_{s}$. To examine the expected reduction in diversity as a function of the number of selected sites ($L$), we ran simulations over a range of values of $L$, up to a maximum of 5x$10^{6}$ (Figure 3). To obtain an estimate of $L$, we calculated the approximate likelihood of our observed data based on the proportion of simulations in which synonymous diversity, $\pi_{s}$, was less than 0.0001 from our empirical estimate (Figure 3).
